---
title: "Data normalisation and batch correction"
author: "Jelena Čuklina"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette summarises the key steps for data batch effect diagnostics and correction using `{proBatch}` package.

Since in most cases both data table and sample annotation file require some tuning, 
we advise to go through 'sample_annotation_preparation' and 'proteome_cleaning' vignettes first.

Here we will complete the following:
1. Run diagnostics before that helps to understand if there are batch effect and 
what type of batch effects it is;
2. Normalize the data (correct for the batch effects).
3. Run the diagnostics again to see how the data changed.

```{r}
library(proBatch)
```
the vignette heavily uses `tidyverse` (also the package), so we load it also
```{r}
library(tidyverse)
```

## Configure the run: specify the columns as they are named (or need to be named) in your dataframes
```{r define the configurations}
feature_id_column = 'peptide_group_label'
measure_column = 'Intensity'
sample_id_column = 'FullRunName'
```

Load the sample annotation. Minimally, this is expected to contain:

1. `sample_id_column` - absolutely unique identifier of experimental file

2. biological identifier (can be repeated if biological replicates are available)

3. biological covatiates, in our case these will be:

  a. `Strain`

  b. `Diet`

  c. `Sex`

  d. `Age_Days`

4. Technical covariates, in this set common for MS run

  a. `SacrificeDate` - date when tissues were extracted

  b. `ProteinPrepDate` - date when samples were prepared

  c. `digestion_batch` - in this case, derived from the previous one, but could be a person who prepared the sample (also multiple variables describing changes in the sample preparation reagents, etc can be reflected)

  d. `RunDate` (and `RunTime`, if available) - will be used to determine run order

  e. `MS_batch` - number of MS batches, sometimes can be inferred from `RunDateTime`, as shown below

  f. in future, this package will also factor in the instrument. In this case, all samples were run on the same instrument
```{r sample_annotation}
data("example_sample_annotation")
biological_covariates = c('Strain', 'Diet', 'Sex', 'Age_Days')
technical_covariates = c('RunDate', 'RunTime', 'SacrificeDate','ProteinPrepDate')
biospecimen_id_col = 'EarTag'
sample_id_column = 'FullRunName'
```
Check the color scheme:
```{r check_color_scheme}
  nrow_prev = floor(sqrt(length(color_list)))
par(mfrow = c(nrow_prev, ceiling(length(color_list)/nrow_prev)),
     oma = c(5,4,1,1) + 0.1,
     mar = c(.51,.51,.51,.51) + 0.1)
for(i in 1:length(color_list)) {scales::show_col(color_list[[i]]); title(main = names(color_list[i]))}
```

Adjust, if necessary:
```{r adjust_colors}
names_sex = names(color_list$Sex)
color_list$Sex = c('pink', 'darkblue')
names(color_list$Sex) = names_sex
names_MS_batch = names(color_list$MS_batch)
color_list$MS_batch = scales::brewer_pal(type = 'qual', palette = 'Dark2')(length(color_list$MS_batch))
#Don't forget to rename (otherwise pheatmap won't work!)
names(color_list$MS_batch) <- names_MS_batch
```

```{r check_colors_again}
  nrow_prev = floor(sqrt(length(color_list)))
par(mfrow = c(nrow_prev, ceiling(length(color_list)/nrow_prev)),
     oma = c(1,1,.51,1) + 0.1,
     mar = c(.51,.51,.51,.51) + 0.1)
for(i in 1:length(color_list)) {scales::show_col(color_list[[i]]); title(main = names(color_list[i]))}
```

1. First, we need to prepare colors, so that they are the same across the plots:
```{r create_colors}
columns_for_plotting = c("Strain", "Diet","Sex","Age_Days",
                         "MS_batch_relaxed", "MS_batch", 
                         "RunDateTime", "ProteinPrepDate", "SacrificeDate")
annotation_for_plotting = sample_annotation_to_colors(sample_annotation, 
                                         columns_for_plotting = columns_for_plotting,
                                        sample_id_column = 'FullRunName',
                                        #if a certain "numeric" column like prep date is actually a factor, specify it here
                                        factor_columns = "ProteinPrepDate", 
                                        #and the other way around, this is still in development
                                        not_factor_columns = "Age_Days", 
                                        #if you have >100 samples, you might want to merge 'singletons' to see the repeated ones
                                        rare_categories_to_other = T, 
                                        numerics_to_log = F)
color_list = annotation_for_plotting$list_of_colors
df_for_annotation = annotation_for_plotting$sample_annotation
```


To see why it matters, let's load the proteome (cleaned as described in {proteome_cleaning} vignette)
Load the protome
```{r load_proteome}
data("example_proteome")
#TODO:
#data('clean_proteome')
```

Few things should be checked first:
1. Average signal intensity vs sample injection order, individually for each instrument
2. Boxplots to examine the differences in the distribution 
3. Sample heatmap (with or without clustering, in the latter case, with different order)
4. iRT peptide and/or spike-in patterns

#1. Plot average signal intensity
```{r plot_mean}
plot_sample_mean(log_transformed_proteome, sample_annotation,
                             sample_id_col = 'FullRunName',
                             order_column = 'RunDateTime',
                             batch_column = NULL,
                             color_by_batch = F, theme = 'classic',
                             title = NULL, color_scheme = 'brewer')
```

As the previous graph suggests, there are batch effects between the machine cleaning.
Even if you don't have access to the machine cleaning procedures, but only the timestamps of the files, these can still be inferred (see the {sample_annotatation_preparation} vignette)







```{r plot_heatmap, fig.width=15, fig.height=20}
library(pheatmap)
test = log_transformed_proteome
test[is.na(test)] = 0
pheatmap(test, cluster_rows = F, cluster_cols = F, 
         color = c('black', colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100)), 
         annotation_col = df_for_annotation, annotation_colors = color_list, 
         show_rownames = F, show_colnames = F)

```
Repeat heatmap with requants removed:
Remove the requant values (values with m_score of 2)
```{r}
example_prot_no_requants = example_proteome %>% filter(m_score <= 1)
```

```{r}
proteome_matrix_no_requants = convert_to_matrix(example_prot_no_requants,
                                    feature_id_column = 'peptide_group_label',
                                    measure_column = 'Intensity',
                                    sample_id_column = 'FullRunName')

log_transformed_proteome_n_rqnts = log2(proteome_matrix_no_requants +1)
```

```{r plot_heatmap_2, fig.width=15, fig.height=20}
test_nrqnt = log_transformed_proteome_n_rqnts
test_nrqnt[is.na(test_nrqnt)] = 0
pheatmap(test_nrqnt, cluster_cols = F, 
         color = c('black', colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100)), 
         annotation_col = df_for_annotation, annotation_colors = color_list, 
         show_rownames = F, show_colnames = F, main = 'no requants')
```

Repeat the signal intensity colored by batch:
```{r}
plot_sample_mean(log_transformed_proteome, sample_annotation,
                             sample_id_col = 'FullRunName',
                             order_column = 'RunDateTime',
                             batch_column = 'MS_batch_relaxed',
                             color_by_batch = T, theme = 'classic',
                             title = NULL, color_scheme = color_list[['MS_batch_relaxed']])
```

```{r}
plot_sample_mean(log_transformed_proteome, sample_annotation,
                             sample_id_col = 'FullRunName',
                             order_column = 'RunDateTime',
                             batch_column = batch_column,
                             color_by_batch = T, theme = 'classic',
                             title = NULL, color_scheme = color_list[['MS_batch']])
```
Now let's clean the data: 
we allow not more than 30% missing values per batch (`threshold_batch`) and not more than
50% of missing values total (`threshold_global`). 
More filters of this kind see in a package `SWATH2stats`
```{r}
print('number of peptides before cleaning')
print(nrow(example_proteome))
proteome_clean = clean_requants(example_proteome, sample_annotation,
                                batch_column = 'MS_batch',
                                feature_id_column = 'peptide_group_label',
                                threshold_batch = .3, threshold_global = .5)
print('number of peptides after cleaning')
print(nrow(proteome_clean))
```

#as ComBat will choke if there are peptides missing in the whole batch, we remove them now
```{r}
proteome_final = remove_peptides_with_missing_batch(proteome_clean,
                                               batch_column = 'MS_batch',
                                               feature_id_column = 'peptide_group_label')
print('number of peptides after missing batch removed peptides')
print(nrow(proteome_final))
```

Now let's convert the matrix for final use:
```{r}
proteome_matrix = convert_to_matrix(proteome_clean,
                                    feature_id_column = 'peptide_group_label',
                                    measure_column = 'Intensity',
                                    sample_id_column = 'FullRunName')
```

### Normalization:
1. Log-transform the data:
```{r}
log_transformed_proteome = log2(proteome_matrix +1)
```

```{r plot_heatmap_3, fig.width=15, fig.height=20}
test_cleaned = log_transformed_proteome
test_cleaned[is.na(test_cleaned)] = 0
pheatmap(test_cleaned, cluster_rows = F, cluster_cols = F, 
         color = c('black', colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100)), 
         annotation_col = df_for_annotation, annotation_colors = color_list, 
         show_rownames = F, show_colnames = F, main = 'no requants, cleaned in each batch')
```

#as ComBat will choke if there are peptides missing in the whole batch, we remove them now
```{r}
proteome_final = remove_peptides_with_missing_batch(proteome_clean,
                                               batch_column = 'MS_batch',
                                               feature_id_column = 'peptide_group_label')
print('number of peptides after missing batch removed peptides')
print(nrow(proteome_final))
```

Now let's convert the matrix for final use:
```{r}
proteome_matrix = convert_to_matrix(proteome_final,
                                    feature_id_column = 'peptide_group_label',
                                    measure_column = 'Intensity',
                                    sample_id_column = 'FullRunName')
log_transformed_proteome = log2(proteome_matrix +1)
```

```{r plot_heatmap_4, fig.width=15, fig.height=20}
test_final = log_transformed_proteome
test_final[is.na(test_final)] = 0
pheatmap(test_final, cluster_cols = F, 
         color = c('black', colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100)), 
         annotation_col = df_for_annotation, annotation_colors = color_list, 
         show_rownames = F, show_colnames = F, main = 'no requants, cleaned across batches')
```

```{r boxplot, fig.width=15, fig.height=20}
gg_boxplot(proteome_final %>% mutate(logInt = log2(Intensity +1)), df_for_annotation, batch_column = 'MS_batch', measure_col = 'logInt', order_column = 'RunDateTime', fill_batch = T)
```


### Normalization:
1. Log-transform the data:
```{r}
log_transformed_proteome = log2(proteome_matrix +1)
```

```{r}
q_norm_proteome = quantile_normalize(log_transformed_proteome)
```

We can check the boxplot after the quantile, quite obvious it looks well
```{r}
q_norm_long = matrix_to_long(q_norm_proteome, sample_annotation, measure_col = 'logInt', step = 'quantile normalized')
gg_boxplot(q_norm_long, df_for_annotation, batch_column = 'MS_batch', measure_col = 'logInt', order_column = 'RunDateTime', fill_batch = T)
```

Also, the heatmap
```{r}
test_q_norm = q_norm_proteome
test_q_norm[is.na(test_q_norm)] = 0
pheatmap(test_q_norm, cluster_cols = F, 
         color = c('black', colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100)), 
         annotation_col = df_for_annotation, annotation_colors = color_list, 
         show_rownames = F, show_colnames = F, main = 'heatmap after quantile normalization')
```

Let's check iRT peptides:
```{r}
data("example_peptide_annotation")
plot_iRTs(q_norm_long %>% merge(example_peptide_annotation, by.x = feature_id_column, by.y = 'Peptide'), sample_annotation, batch_column = batch_column, measurement.col = 'logInt', order_column = 'order')
```

*Caution*: for non-linear fit, using order rather than physical run time is recommended
```{r warnings = FALSE}
		loess_norm_proteome =  normalize_custom_fit(q_norm_proteome, sample_annotation,
		                                           batch_col = 'MS_batch',
		                                           feature_id_col = 'peptide_group_label',
		                                           sample_id_col = 'FullRunName',
		                                           measure_col = 'Intensity',
		                                           sample_order_col = 'order',
		                                           fit_func = fit_nonlinear,
		                                           return_long = T,
		                                           fitFunc = 'loess_regression', with_df = F)
		loess_norm_proteome_matrix = convert_to_matrix(loess_norm_proteome,
		                                    feature_id_column = 'peptide_group_label',
		                                    measure_column = 'Intensity_normalized',
		                                    sample_id_column = 'FullRunName')
```

Demonstrate how it worked on the example of the peptides of one protein:
```{r}
test_protein = "1/sp|Q78JT3"
test_peptides = unique(proteome_final$peptide_group_label[proteome_final$ProteinName == test_protein])[1:6]
```

Before the correction
```{r plot_Haao_peptides}
gg = plot_peptide_level(pep_name = test_peptides, data_df_long = loess_norm_proteome, 
                   sample_annotation = sample_annotation, sample_id_column = sample_id_column,
                   batch_column = 'MS_batch', feature_id_column = feature_id_column, 
                   measurement.col = 'Intensity', order_column = 'order', 
                   title = 'Peptides of Haao protein before the LOESS fit', facet_by_batch = F)
gg
```


```{r plot_Haao_after_loess}
gg = plot_peptide_level(pep_name = test_peptides, data_df_long = loess_norm_proteome, 
                   sample_annotation = sample_annotation, sample_id_column = sample_id_column,
                   batch_column = 'MS_batch', feature_id_column = feature_id_column, 
                   measurement.col = 'Intensity_normalized', order_column = 'order', 
                   title = 'Peptides of Haao protein after the LOESS fit', facet_by_batch = F)
gg

```

To compare before and after the fit, we need to join the data matrices first
```{r plot_Haao_with_the_fit}
joined_df = join_data_matrices(matrix_list = list(log_transformed_proteome,
                                           q_norm_proteome, 
                                           loess_norm_proteome_matrix),
                               sample_annotation = sample_annotation,
                               Step = c('1_raw_logTransformed',
                                        '2_quantile_normalized',
                                        '3_loess_normalized'))
fit_df = loess_norm_proteome %>% select(peptide_group_label, FullRunName, fit, MS_batch)
gg = plot_with_fitting_curve(pep_name = test_peptides, fit_df = fit_df, 
                             sample_annotation = sample_annotation, 
                             data_df_all_steps = joined_df, batch_column = 'MS_batch', 
                             order_column = 'order', measurement.col = measure_column,
                             fit_value_var = 'fit', fit_step = '2_quantile_normalized',
                             color_by_batch = color_list$MS_batch, title = 'test')
gg
```


4. Finally, ComBat to equalize the means of each peptide:
Before, we need to filter for peptides that are detected in every batch

```{r ComBat}
		comBat_norm_proteome = correct_with_ComBat(loess_norm_proteome_matrix,
		                                           sample_annotation = sample_annotation,
		                                           batch_column = 'MS_batch',
		                                           par.prior = F)

```

### Diagnostics

```{r color_list_to_df}
color_df = color_list_to_df(color_list, df_for_annotation)
cluster_samples(log_transformed_proteome, color_df, plot_title = '1_raw_log_transformed')

```

Examine the replicates:
```{r replicates}
earTags = c("ET1420", "ET1322", "ET2154", "ET1518", "ET1515", "ET1906", "ET1506",
            "ET2078", "ET1566", "ET1354", "ET1585", "ET1681", "ET1524")
samples_to_examine = sample_annotation %>%
  filter(EarTag %in% earTags) %>% pull(FullRunName)
data_matrix = log_transformed_proteome[,samples_to_examine]
p = plot_sample_corr_distribution(data_matrix = log_transformed_proteome, sample_annotation, repeated_samples = samples_to_examine, sample_id_col = 'FullRunName', batch_col = 'MS_batch', plot_param = 'batch_replicate')
p
```

Visualise principal components 1 and 3:
```{r}
plot_pca(log_transformed_proteome, df_for_annotation, factor_to_color = 'MS_batch', 
         colors_for_factor = color_list$MS_batch, PC_to_plot = c(1, 3))
```

Quantify your batch effects with PVCA (will take a while if you have a large data matrix)
```{r}

```


